Staged Abstract Interpreters

### 2018-11-13 paper/10_intro.tex
fix typos and restructure

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close connection between 
compilers and interpreters. 
The first Futamura projection specifically shows that specializing an interpreter with respect to the input 
program yields an equivalent executable.
Partial evaluation \cite{DBLP:books/daglib/0072559} was the first proposed approach to realize Futamura projections,
which first identifies the binding-time of variables in the program, i.e., they can be known statically or dynamically, 
and then specializes the program with respect to the given input of program.
However, precisely analyzing binding-time for an arbitray program is hard in general. As an alternative and programmatic
approach to specialization and partial evaluation, multi-stage programming (MSP) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} 
requires the stage annotations (i.e., binding-time annotations) from the programmers, these staging annotations 
control which part of the input program should be specialized.
As a simple example, we consider a power function: 
\begin{lstlisting}
def power(b: Int, e: Int): Int = 
  if (e == 0) 1 else e * power(b, e-1)
\end{lstlisting}
If \texttt{e} is statically known, say 5, then we can specialize \texttt{power} with respect to \texttt{e = 5} and 
generate a new specialized function as following:
\begin{lstlisting}
def power5(b: Int) = b * b * b * b * b
\end{lstlisting}

@@

Multi-stage programming has been widely used to improve the performance in many domains, 
such as optimizing compilers and domain-specific languages \cite{DBLP:conf/pldi/RompfSBLCO14, DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13, DBLP:journals/jfp/CaretteKS09},
numerical computation \cite{PGL-038, DBLP:conf/pepm/AktemurKKS13}, 
generic programming \cite{DBLP:journals/pacmpl/Yallop17, Ofenbeck:2017:SGP:3136040.3136060}, 
data processing \cite{DBLP:conf/oopsla/JonnalageddaCSRO14, DBLP:conf/popl/KiselyovBPS17}, 
query compilation in databases \cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18},
etc.
Likewise, specializing static analyses by partial evaluation had been emerged in late 90s 
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587, ashley:practical}, 
and indeed it is able to effectively remove the interpretive overhead of 
repeatedly traversing the abstract syntax tree. 
However, despite of the previous work are mostly focus on one particular analysis,
it is worth to investigate the idea from a modern perspective based on meta-programming,
especially for a general abstract interpreter that models direct-style $\lambda$-calculus, imperative 
features such as mutation, and different abstract domains.
One technical challenge is the binding-time engineering when non-determinism, 
fixed-point iterations, different abstract domains and staging are introduced at the same time.
In this paper, we present an end-to-end design and implementation of staging an abstract interpreter,
that means ont only the abstract interpreter, but also the data structure of abstract domains, 
abstract environment and heap, and the fixpoint iteration are staged.

@@

When staging a concrete interpreter, the programmers need to distinguish 
static and dynamic values --- the given program to be executed by the 
interpreter is classified as static because it is known at compile-time, 
and the inputs to that program are dynamic.  However, when staging an 
abstract interpreter, this distinction does not exist anymore. Because the 
abstract interpreter instantiates all the inputs as some form of abstract 
values, which are usually top elements in their abstract domains and are 
also statically known. 
Then what is the point of staging if there is no such distinction?
\note{TR: I don't understand this. The program
structure is static, the abstract values are still dynamic, no? They 
change in every iteration of the fixpoint algorithm}
A surprising by-product of thinking about this question is to realize that 
we can apply the staged abstract interpreter on \textit{open} programs, and
the free variables representing other parts of the program (e.g., libraries) 
are dynamic input, therefore we obtain a modular analysis through staging, 
mechanically.

The use of staged abstract interpreters on open programs improves the scalability. 
\note{TR: why is it a big deal. On Closed programs we obtain a constant factor, but modular analysis has different asymptotics}
One challenge in static analysis of modern software is that the programs are usually 
shipped with large library code, for example, it has been shown that analyzing a simple
"Hello World" program in Java depends on additional 3,000 classes in the library \cite{DBLP:conf/oopsla/KulkarniMZN16}.
Soundly and precisely analyzing these libraries sometimes are expensive and unnecessary repeatedly.
When applying the staged abstract interpreter on libraries, we leave the unknown 
calling context as dynamic value and generate partial analyzing result which is 
repsented as a residul program. The partial analyzing result can be resued and composed
with the analyzing result of application code when available later. 
It has been observed that partially applying context-sensitivity on selected portion of the program could
improve the precision and efficiency \cite{zipper2018, Kastrinis:2013:HCP:2491956.2462191}. 
We show that staging abstract interpreters as an approach to effectively implement hybird context-sensitivity\todo{}.

@@

\begin{itemize}
\item Starting from a definitional abstract interpreter for a higher-order functional language,
  whose syntax and semantics are described in Section~\ref{bg_lang},
  we show how to turn it into a staged abstract interpreter (SAI) by adding stage annotations. 
  The design and implementation of SAI using Scala and Lightweight Modular Staging (LMS) framework is 
  described in Section~\ref{sai}.
  We later extend the staged abstract interpreter to support imperative language features 
  (Section~\ref{cases_imp}), such as states, loops and exceptions.
\item We demonstrate that the staged abstract interpreter not only improves the \textit{efficiency}
  but also the \textit{scalability}, which are two major issues in static analysis, if we apply it on open
  programs. \todo{more on this point.}
\item To demonstrate the feasibility and applicability of the proposed approach, we conduct three case studies in Section~\ref{cases_study}. 
  First we revisit Abstract Compilation \cite{Boucher:1996:ACN:647473.727587}, which is a closure generation 
  technique applied on 0-CFA. We show that it can be understood and easily implemented as an instance of 
  our staged abstract interpreter. 
  Second, we demonstrate sophisticated control-flow analysis techniques, such as adaptive context/flow/path-sensitive 
  analysis and store-widening can be integrated with staged abstract interpreters without losing any precision.
  Finally, we show that in an imperative language, the idea can be applied to an abstract interpreter with
  an abstract domain that performs numerical analysis.
  % (Optional: taint/information flow analysis?)
\item We empirically evaluate the performance improvement by staging an big-step abstract interpreter 
  for control-flow analysis. We compare against with the unstaged version, a small-step version (unstaged), 
  and an optimizing small-step version.
\end{itemize}

>>>>

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close connection between 
compilers and interpreters. 
The first Futamura projection specifically shows that specializing an interpreter with respect to the input 
program yields an equivalent executable.
Partial evaluation \cite{DBLP:books/daglib/0072559} was the first proposed approach to realize Futamura projections,
which first identifies the binding-time of variables in the program, i.e., they can be known statically or dynamically, 
and then specializes the program with respect to the given input of program.
However, precisely analyzing binding-time for an arbitray program is hard in general. As an alternative and programmatic
approach to specialization and partial evaluation, multi-stage programming (MSP) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} 
requires the stage annotations (i.e., binding-time annotations) from the programmers. These staging annotations 
control which part of the input program should be specialized.
As a simple example, we consider a power function: 
\begin{lstlisting}
def power(b: Int, e: Int): Int = 
  if (e == 0) 1 else e * power(b, e-1)
\end{lstlisting}
If \texttt{e} is identified as statically known by the programmer, say 5, then we can specialize \texttt{power} 
with respect to \texttt{e = 5} and generate a new specialized function as following:
\begin{lstlisting}
def power5(b: Int) = b * b * b * b * b
\end{lstlisting}

@@

Multi-stage programming has been widely used to improve the performance in many domains, 
such as optimizing compilers and domain-specific languages \cite{DBLP:conf/pldi/RompfSBLCO14, DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13, DBLP:journals/jfp/CaretteKS09},
numerical computation \cite{PGL-038, DBLP:conf/pepm/AktemurKKS13}, 
generic programming \cite{DBLP:journals/pacmpl/Yallop17, Ofenbeck:2017:SGP:3136040.3136060}, 
data processing \cite{DBLP:conf/oopsla/JonnalageddaCSRO14, DBLP:conf/popl/KiselyovBPS17}, 
query compilation in databases \cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18},
etc.
Likewise, specializing static analyses by partial evaluation emerged in late 90s 
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587, ashley:practical}, 
and indeed it is able to effectively remove the interpretive overhead of 
repeatedly traversing the abstract syntax tree. 
However, despite of the previous works mostly focused on one particular analysis,
or required to rewrite the analyzer,
it is worth to investigate the idea from a modern perspective based on meta-programming,
especially for a general abstract interpreter that models direct-style $\lambda$-calculus, imperative 
features such as mutation, and different abstract domains.
One technical challenge is the binding-time engineering when non-determinism, 
fixed-point iterations, different abstract domains and staging are introduced at the same time.
In this paper, we present an end-to-end design and implementation of staging an abstract interpreter,
that means not only the interpreter that traverses the abstract syntax tree, 
but also the data structure of abstract domains, abstract environment and heap, 
and the fixpoint iteration are staged.

@@

When staging a concrete interpreter, the programmers need to distinguish 
static and dynamic values --- the given program to be executed by the 
interpreter is classified as static because it is known at compile-time, 
and the inputs to that program are dynamic.  However, when staging an 
abstract interpreter, this distinction does not exist anymore. Because the 
abstract interpreter instantiates all the inputs as some form of abstract 
values, which are usually top elements in their abstract domains and are 
also statically known. 
Then what is the point of staging if there is no such distinction?
\note{TR: I don't understand this. The program
structure is static, the abstract values are still dynamic, no? They 
change in every iteration of the fixpoint algorithm}
A surprising by-product of thinking about this question is to realize that 
we can apply the staged abstract interpreter on \textit{open} programs, and
the free variables representing other parts of the program (e.g., libraries) 
are dynamic inputs, therefore we obtain a modular analysis through staging, 
mechanically.

The use of staged abstract interpreters on open programs improves the scalability. 
\note{TR: why is it a big deal. On Closed programs we obtain a constant factor, but modular analysis has different asymptotics}
One challenge in static analysis of modern software is that the programs are usually 
shipped with large library code, for example, it has been shown that analyzing a simple
"Hello World" program in Java depends on additional 3,000 classes in the library \cite{DBLP:conf/oopsla/KulkarniMZN16}.
Soundly and precisely analyzing these libraries sometimes are expensive and unnecessarily repeated.
When applying the staged abstract interpreter on libraries, we leave the unknown 
calling context as dynamic value and generate partial analyzing result which is 
repsented as a residul program. The partial analyzing result can be resued and composed
with the analyzing result of application code when available later. 
It has been observed that partially applying context-sensitivity on selected portion of the program could
improve the precision and efficiency \cite{zipper2018, Kastrinis:2013:HCP:2491956.2462191}. 
We show that staging abstract interpreters as an approach to effectively implement hybird context-sensitivity\todo{}.

@@

\begin{itemize}
\item Starting from a definitional abstract interpreter for a higher-order functional language,
  whose syntax and semantics are described in Section~\ref{bg_lang},
  we show how to turn it into a staged abstract interpreter (SAI) by adding stage annotations. 
  The design and implementation of SAI using Scala and Lightweight Modular Staging (LMS) framework 
  are described in Section~\ref{sai}.
  We later extend the staged abstract interpreter to support imperative language features 
  (Section~\ref{cases_imp}), such as states, loops and exceptions.
\item We demonstrate that the staged abstract interpreter not only improves the \textit{efficiency}
  but also the \textit{scalability}, which are two major issues in static analysis, if we apply it on open
  programs. \todo{more on this point.}
\item To demonstrate the feasibility and applicability of the proposed approach, we conduct three case studies in Section~\ref{cases_study}. 
  First, we revisit Abstract Compilation \cite{Boucher:1996:ACN:647473.727587}, which is a closure generation 
  technique applied on 0-CFA. We show that it can be understood and easily implemented as an instance of 
  our staged abstract interpreter. 
  Second, we demonstrate sophisticated control-flow analysis techniques, such as adaptive context/flow/path-sensitive 
  analysis and store-widening can be integrated with staged abstract interpreters without losing any precision.
  Finally, we show that in an imperative language, the idea can be applied to an abstract interpreter with
  an abstract domain that performs numerical analysis.
  % (Optional: taint/information flow analysis?)
\item We empirically evaluate the performance improvement by staging an big-step abstract interpreter 
  for control-flow analysis. We compare against with the unstaged version, a small-step version (unstaged), 
  and an optimizing small-step version.
\end{itemize}

### 2018-11-13 paper/20_bg.tex
fix typos and restructure

\section{Background}

In this section, we first described the language we will model and analyze, and present an unstaged concrete 
interpreter for it. Then, starting from this unstaged concrete interpreter, we proceed in two directions:
by abstracting it we obtain an unstaged abstract interpreter; by staging it we obtain a staged concrete
interpreter.

>>>>

\section{Preliminary}

In this section, we first describe the abstract syntax of the language we will analyze, then present an 
generic interface of interpreters, and instantiate an unstaged concrete interpreter.
Then, starting from this unstaged concrete interpreter, we proceed in two directions:
by abstracting it we obtain an unstaged abstract interpreter (Section~\ref{unstaged_abs}); 
by staging it we obtain a staged concrete interpreter (Section~\ref{stagedinterp}).

### 2018-11-13 paper/21_bg_interp.tex
fix typos and restructure

\subsection{An Unstaged Concrete Interpreter} \label{bg_lang}

As higher-order functions had been widely introduced in mainstream programming languages (such as Python and Java \todo{cite}), 
in this paper, we consider a tiny higher-order functional language \TLang, which is based on direct-style $\lambda$-calculus
with numbers, arithmetics, recursions and conditionals.
In Section~\ref{cases_imp}, we will see adding more imperative features to the language.
Since we are mostly interested in analyzing the dynamic behavior of the program, the static semantics and
type system will be omitted. We also assume the input programs are well-typed and variables are distinct.
The abstract syntax is described below:

@@

\paragraph{Generic Interface.} 
Before showing the actual implementation of the concrete semantics, let's first see 
a generic interface for the interpreters, which will be used later for both unstaged/staged and concrete/abstract scenarios.
The trait \texttt{Semantics} first declares several abstract type members, such as \texttt{Addr},
\texttt{Value}, \texttt{Env}, \texttt{Store} and returned type \texttt{Ans}. 
Initial values for environments and stores, and few abstract methods that manipulate environments and stores are also declared.

@@

The implementation of the skeleton interpreter traverses the abstract syntax tree and should be straightforward.
It is worth to note that the interpreter is written in open-recursive style -- 
it can not refer to itself directly, instead, \texttt{eval} takes an additional 
parameter \texttt{ev} referring to itself; this allows us to instrument the interpreter
from outside. Accordingly, the function \texttt{close} for lifting lambda term to closure values
is also written in this style.
We define a default implementation \texttt{fix} to close the recursion by recursively use the result 
of \texttt{fix(ev)} for \texttt{ev}. 
Finally, the top-level evaluation function \texttt{eval\_top} is defined.

\paragraph{Stage Polymorphism.}
Readers may notice an abstract higher-kinded type \texttt{R[+T]} is defined in the trait, that is a
type-level function and will be used to specify the binding-time, i.e., staged or not; the annotation 
\texttt{+} indicates type parameter \texttt{T} is covariant.
The LMS framework we use here allows us to specify binding-time purely based on types.
All occurrences of type \texttt{Env}, \texttt{Store}, etc. are wrapped with \texttt{R} 
essentially because they may potentially being dynamic when staging is introduced.
Obviously, the unstaged higher-kinded type then can be simply instantiated as an identity type function: 
\texttt{type R[+T] = T}. In Section~\ref{stagedinterp}, \texttt{R} will be instantiated 
as a staging annotation. \todo{where talk about stage poly?}

\paragraph{Concrete Instantiation.}
Now we can instantiate a concrete interpreter. We first concretize the abstract types as follows.
Since the interpreter is concrete, for the propose that properly models the heap, the type \texttt{Int} 
is used for addresses space \texttt{Addr}. The environment and store are just ordinary \texttt{Map} in Scala.
The values can be either a tagged number \texttt{NumV} or a closure \texttt{CloV} that contains a lambda term
and an environment.

@@

\texttt{get} and \texttt{put} directly manipulate the map of environment and store. 
An allocation function is used when we need to update the store: 
in the concrete case, \texttt{alloc} always returns a fresh address of the store.
\texttt{close} and \texttt{num} lift syntactic literals to values.
\texttt{apply\_closure} takes the function value, argument value and the latest store,
then extends the environment and store after allocating a fresh address, and continue
evaluates the body \texttt{e} of the lambda term.

>>>>

\subsection{Abstract Syntax} \label{bg_lang}

As higher-order functions had been widely introduced to mainstream programming languages (such as Python and Java \todo{cite}), 
in this paper, we consider a tiny higher-order functional language \TLang, which is based on direct-style $\lambda$-calculus
with numbers, arithmetics, recursions and conditionals.
In Section~\ref{cases_imp}, we will see adding more imperative features to the language.
Since we are mostly interested in analyzing the dynamic behavior of the program, the static semantics and
type system will be omitted. We also assume the input programs are well-typed and variables are distinct.
The abstract syntax is described below:

@@

\subsection{Generic Interface} 
Before showing the actual implementation of the concrete semantics, let's first see 
a generic interface for the interpreters, which is semantics-agnostic and can be 
"interpreted" by either concrete or abstract, unstaged or staged semantics by our choice.
The trait \texttt{Semantics} first declares several abstract type members, such as \texttt{Addr},
\texttt{Value}, \texttt{Env}, \texttt{Store} and returned type \texttt{Ans}. 
Initial values for environments and stores, and few abstract methods that manipulate environments and stores are also declared.

@@

The implementation of the skeleton interpreter traverses the abstract syntax tree and should be straightforward.
It is worth noting that the interpreter is written in open-recursive style -- 
it can not refer to itself directly, instead, \texttt{eval} takes an additional 
parameter \texttt{ev} referring to itself; this allows us to instrument the interpreter
from outside. Accordingly, the function \texttt{close} for lifting lambda term to closure values
is also written in this style.
We define a default implementation \texttt{fix} to close the recursion by recursively using the result 
of \texttt{fix(ev)} for \texttt{ev}. 
Finally, the top-level evaluation function \texttt{eval\_top} is defined.

\paragraph{Stage Polymorphism.}
Readers may notice an abstract higher-kinded type \texttt{R[+T]} is defined in the trait, that is a
type-level function and will be used to specify the binding-time, i.e., staged or not; the annotation 
\texttt{+} indicates type parameter \texttt{T} is covariant.
The LMS framework we use here allows us to specify binding-time purely based on types.
All occurrences of type \texttt{Env}, \texttt{Store}, etc. are wrapped with \texttt{R} 
essentially because they may potentially be dynamic when staging is introduced.
Obviously, the unstaged higher-kinded type then can be simply instantiated as an identity type function: 
\texttt{type R[+T] = T}. In Section~\ref{stagedinterp}, \texttt{R} will be instantiated 
as a staging annotation. \todo{where talk about stage poly?}

\subsection{Concrete Instantiation}
Now we can instantiate a concrete interpreter. We first concretize the abstract types as follows.
Since the interpreter is concrete, for the propose that properly models the heap, the type \texttt{Int} 
is used for address space \texttt{Addr}. The environment and store are just ordinary \texttt{Map} in Scala.
The values can be either a tagged number \texttt{NumV} or a closure \texttt{CloV} that contains a lambda term
and an environment.

@@

\texttt{get} and \texttt{put} directly manipulate the map of environment and store. 
An allocation function is used when we need to update the store: 
in the concrete case, \texttt{alloc} always returns a fresh address of the store.
\texttt{close} and \texttt{num} lift syntactic literals to values.
\texttt{apply\_closure} takes the function value, argument value and the latest store,
then extends the environment and store after allocating a fresh address, and continue
evaluating the body \texttt{e} of the lambda term.

### 2018-11-13 paper/22_bg_abs.tex
fix typos and restructure

\subsection{From Interpreter to Abstract Interpreter} \label{unstaged_abs}

@@

\paragraph{Stage Polymorphic Lattices.}
To effectively reuse the code between unstaged and staged part, we also make the lattice structure to
be stage polymorphic. A lattice type class parameterizes over an element type \texttt{E} and defines
five operations for \texttt{R[E]}, such as meet, join, and ordering relation; again, \texttt{R} 
is the higher-kinded type indicating the binding-time.
\begin{lstlisting}
trait Lattice[E, R[+_]] {
  val bot: R[E];  val top: R[E]
  def $\sqsubseteq$(l1: R[E], l2: R[E]): R[Boolean]
  def $\sqcup$(l1: R[E], l2: R[E]): R[E]
  def $\sqcap$(l1: R[E], l2: R[E]): R[E]
}
\end{lstlisting}

@@

\paragraph{Abstract Semantics.}
The operations @get@ and @put@ on abstract stores are slightly changed according to
the abstract semantics: @get@ uses the bottom element in the @Value@ lattice if the 
queried address not exists; @put@ performs the update by joining with the existing values.
@alloc@ simply invokes the @Addr@ constructor.
\texttt{close} and \texttt{num} lift syntactic literals to our abstract domain, i.e., 
a singleton set that contains only one \texttt{CloV} or \texttt{NumV} object.

@@

\paragraph{Fixpoint Iteration.}
We described the abstract semantics modularly in the last section, however, the abstract 
interpreter may not terminate for some input program. In this section, we use a memoization
technique to ensure its termination. This technique is also widely used in the name of co-inductive caching algorithm
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800} or truncated depth-first evaluation 
\cite{Rosendahl:AbsIntPL}.

@@

Finally, we override the definition of @eval_top@ by instantiating @CacheFix@ with @eval@ and start 
the first iteration.

>>>>

\section{From Interpreters to Abstract Interpreters} \label{unstaged_abs}

@@

\subsection{Stage Polymorphic Lattices} \label{stagedpoly_lat}
To effectively reuse the code between unstaged and staged part, we also make the lattice structure to
be stage polymorphic. A lattice type class parameterizes over an element type \texttt{E} and defines
five operations for \texttt{R[E]}, such as meet, join, and ordering relation; again, \texttt{R} 
is the higher-kinded type indicating the binding-time.
\begin{lstlisting}
trait Lattice[E, R[+_]] {
  val bot: R[E];  val top: R[E]
  def $\sqsubseteq$(l1: R[E], l2: R[E]): R[Boolean]
  def $\sqcup$(l1: R[E], l2: R[E]): R[E]
  def $\sqcap$(l1: R[E], l2: R[E]): R[E]
}
\end{lstlisting}

@@

\subsection{Abstract Semantics}
The operations @get@ and @put@ on abstract stores are slightly changed according to
the abstract semantics: @get@ uses the bottom element in the @Value@ lattice if the 
queried address not exists; @put@ performs the update by joining with the existing values.
@alloc@ simply invokes the @Addr@ constructor.
\texttt{close} and \texttt{num} lift syntactic literals to our abstract domain, i.e., 
a singleton set that contains only one \texttt{CloV} or \texttt{NumV} object.

@@

\subsection{Fixpoint Iteration}
We described the abstract semantics modularly in the last section, however, the abstract 
interpreter may not terminate for some input program. In this section, we use a memoization
technique to ensure its termination. This technique is also widely used in the name of co-inductive caching algorithm
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800} or truncated depth-first evaluation 
\cite{Rosendahl:AbsIntPL}.

@@

Finally, we override the definition of @eval_top@ by instantiating @CacheFix@ with @eval@ and starting 
the first iteration.

### 2018-11-13 paper/23_bg_stage.tex
fix typos and restructure

\subsection{From Interpreter to Staged Interpreter} \label{stagedinterp}

\paragraph{Multi-Stage Programming in LMS.}
Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a multi-stage programming framework implemented 
as a Scala library that enables dynamic code generation in a type-safe manner.
Different from the syntactic approach such as MetaOCaml \cite{DBLP:conf/flops/Kiselyov14} that uses quotations, 
LMS distinguishes binding-time solely based on types.
LMS provides a type constructor @Rep[T]@ where @T@ can be an arbitrary type, indicating an expression will be 
known at next stage. All operations act on a @Rep@ expression will be resulted in the generated code. However,
we need to provide intermediate representation and code generation support for type @T@ to LMS. 
Fortunately, the LMS framework already provides
such support for primitive types and commonly used data structure such as arrays and maps; implementing such support
for custom class is also straightforward. Let's go back and see how LMS can be used to specialize the power function
we mentioned in Section 1.

\begin{lstlisting}
val power5 = new DslDriver[Int, Int] {
  def power(b: Rep[Int], x: Int): Rep[Int] =
    if (x == 0) 1 else b * power(b, x-1)
  def snippet(x: Rep[Int]): Rep[Int] = power(x, 5)
}
\end{lstlisting}

In the code shown above, \texttt{power} takes 2 arguments where \texttt{b} is declared as \texttt{Rep[Int]} type 
meaning that \texttt{b} is a representation of \texttt{Int} but whose value will be available at the next stage. 
Meanwhile, the result of the @power@ function will be available at the next stage as will, thus the return value 
of \texttt{power} is also of type \texttt{Rep[Int]}. 
Then we specialize @power@ in function @snippet@ by providing @x@ to @5@.
The generated code for \texttt{power5} is a function that takes only one argument which corresponds to @b@,
and the body of this function multiples @b@ five times.

\paragraph{Staged Concrete Semantics.} In the staging part, we share the same concrete type instantiation, 
i.e. trait \texttt{Concrete}, but we reimplement the staged version for the concrete operations.
Meanwhile, @RepConcInterpOps@ also extends from @LMSOps@ which provides necessary supports for the 
staged version of primitive types and data structures. \todo{what are static, what are dyanmic?}

@@

The first notable change is the abstract type member @R@ is assigned to be @Rep@, and accordingly the affected 
types such as @Env@ and @Store@ become @Rep[Env]@ and @Rep[Store]@. For @num@, we use @lift@ function which is a
built-in in LMS that lifts a current-stage value to next stage.
The interesting thing is how we handle closures and function applications when staging is involved.
In the unstaged version, function @close@ is used to lift a literal lambda term to a closure, which
is a pair of the syntactic lambda term and the enclosing environment.
However, what we desire is the a \textit{compiled} closure, instead of a lifted @CloV@ value like 
what we did for @NumV@ --- in other words, the syntactic term of the lambda expression should be 
eliminated and specialized away.
The specialization of the interpreter with respect to the body of the lambda term proceed under the
assumptions that the argument and latest store will be provided later. \note{to denotational style?, back to HO?}
We will see this is an important observation that enables us to specialize an abstract interpreter modularly.
In the end of @close@, we create a @CompiledClo@ object that contains a staged function, which both will be
exhibited in the generated code.

@@

Then, @apply_closure@ takes three arguments the function value, argument value and the latest store, 
which are both of @Rep@ type. It means basically these is nothing we can do at current stage --- 
so we create a new node @ApplyClosure@ that contains these arguments in the intermediate representation graph,
and in the code generation phase we can emit the code that does the application. \todo{reflectEffect}.

@@

\paragraph{Code Generation.}
When generating code for the next stage, values like @CompiledClo@ and @NumV@ are kept as what it is because
they are intended to be existed in the next stage; however, we need to treat @ApplyClosure@ a little bit 
differently.

>>>>

\section{From Interpreters to Staged Interpreters} \label{stagedinterp}

\subsection{Multi-Stage Programming in LMS}
Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a multi-stage programming framework implemented 
as a Scala library that enables dynamic code generation in a type-safe manner.
Different from the syntactic approach such as MetaOCaml \cite{DBLP:conf/flops/Kiselyov14} that uses quotations, 
LMS distinguishes binding-time solely based on types.
LMS provides a type constructor @Rep[T]@ where @T@ can be an arbitrary type, indicating an expression will be 
known at next stage. All operations acting on a @Rep@ expression will be resulted in the generated code. However,
we need to provide intermediate representation and code generation support for type @T@ to LMS. 
Fortunately, the LMS framework already provides
such support for primitive types and commonly used data structure such as arrays and maps; implementing such support
for custom class is also straightforward. Let's go back and see how LMS can be used to specialize the power function
we mentioned in Section 1.

\begin{lstlisting}
.. = new DslDriver[Int, Int] {
  def power(b: Rep[Int], x: Int): Rep[Int] =
    if (x == 0) 1 else b * power(b, x-1)
  def snippet(x: Rep[Int]): Rep[Int] = power(x, 5)
}
\end{lstlisting}

In the code shown above, \texttt{power} takes 2 arguments where \texttt{b} is declared as \texttt{Rep[Int]} type 
meaning that \texttt{b} is a representation of \texttt{Int} but whose value will be available at the next stage. 
Meanwhile, the result of the @power@ function will be available at the next stage as well, thus the return value 
of \texttt{power} is also of type \texttt{Rep[Int]}. 
Then we specialize @power@ in function @snippet@ by providing @x@ to @5@.
The generated code for \texttt{power5} is a function that takes only one argument which corresponds to @b@,
and the body of this function multiples @b@ five times.

\subsection{Staged Concrete Semantics} 
In the staging part, we share the same concrete type instantiation, 
i.e. trait \texttt{Concrete}, but we reimplement the staged version for the concrete operations.
Meanwhile, @RepConcInterpOps@ also extends from @LMSOps@ which provides necessary supports for the 
staged version of primitive types and data structures. \todo{what are static, what are dyanmic?}

@@

The first notable change is the abstract type member @R@ is assigned to be @Rep@, and accordingly the affected 
types such as @Env@ and @Store@ become @Rep[Env]@ and @Rep[Store]@. For @num@, we use @lift@ function which is a
built-in in LMS that lifts a current-stage value to next stage.
The interesting thing is how we handle closures and function applications when staging is involved.
In the unstaged version, function @close@ is used to lift a literal lambda term to a closure, which
is a pair of the syntactic lambda term and the enclosing environment.
However, what we desire is a \textit{compiled} closure, instead of a lifted @CloV@ value like 
what we did for @NumV@ --- in other words, the syntactic term of the lambda expression should be 
eliminated and specialized away.
The specialization of the interpreter with respect to the body of the lambda term proceed under the
assumptions that the argument and latest store will be provided later. \note{to denotational style?, back to HO?}
We will see this is an important observation that enables us to specialize an abstract interpreter modularly.
In the end of @close@, we create a @CompiledClo@ object that contains a staged function, which all will be
exhibited in the generated code.

@@

Then, @apply_closure@ takes three arguments the function value, argument value and the latest store, 
which are both of @Rep@ type. It means basically there is nothing we can do at current stage --- 
so we create a new node @ApplyClosure@ that contains these arguments in the intermediate representation graph,
and in the code generation phase we can emit the code that does the application. \todo{reflectEffect}.

@@

\paragraph{Code Generation.}
When generating code for the next stage, values like @CompiledClo@ and @NumV@ are kept as what they are because
they are intended to be existed in the next stage; however, we need to treat @ApplyClosure@ a little bit 
differently.

### 2018-11-13 paper/30_sai.tex
fix typos and restructure

\section{Staging Abstract Interpreters} \label{sai}

In the previous section, we have seen an unstaged abstract interpreter and a staged concrete interpreter, 
now we begin describing the implementation of their confluence -- a staged abstract interpreter. 
Unsurprisingly, the staged abstract interpreter has the same abstract semantics with the unstaged version
we presented in Section~\ref{unstaged_abs}.

\subsection{From Abstract Interpreter to Staged Abstract Interpreter}

In this section, we present a principled approach to derive staged abstract interpreter
from its unstaged version. One guiding principle of our approach is the code of abstract semantics
and the code that optimizes should be separated. Thus it is also an advantage of doing staging 
for abstract interpreters: the designer of analysis has no need to rewrite the analysis, and 
the performance improvement comes almost for free, without any sacrifice of soundness or precision.

\paragraph{Staged Lattices.} In Section~\ref{unstaged_abs}, we exploited the higher-kinded type @R@
to leave space for staging lattices, now we instantiate @R@ to @Rep@ and still use powerset as an 
example to present its staged version.

@@

\paragraph{Staged Abstract Semantics.} We have seen how to obtain a staged concrete semantics
based on types, now we take the same approach to obtain a staged abstract semantics.
\todo{what are static, what are dynamic?}
The basic operations largely kept same as the unstaged version, except the types are changed to @Rep@.
Besides, when we update the environment, the identifier @x@ is known statically, but the map of environment
has type @Rep[Map[Ident,Addr]]@, so we apply @lift@ to @x@ to make it as a next-stage value.

@@

\paragraph{Staged Fixpoint Iteration.} 
\todo{staged fixed point iteration with memoization}

@@

Jones-optimal specialization?
Futamura projection?

\subsection{Lattice-theoretic Interpretation}

>>>>

\section{From Abstract Interpreters to Staged Abstract Interpreters} \label{sai}

In the previous section, we have seen an unstaged abstract interpreter and a staged concrete interpreter, 
now we begin describing the implementation of their confluence -- a staged abstract interpreter. 
We present a principled approach to derive staged abstract interpreter
from its unstaged version. One guiding principle of our approach is the code of abstract semantics
and the code that optimizes should be separated. Thus it is also an advantage of doing staging 
for abstract interpreters: the designer of analysis has no need to rewrite the analysis, and 
the performance improvement comes almost for free, without any sacrifice of soundness or precision.
Unsurprisingly, the staged abstract interpreter we present in this section has the same abstract 
semantics with the unstaged version we presented in Section~\ref{unstaged_abs}.

\subsection{Staged Lattices} 
In Section~\ref{stagedpoly_lat}, we exploited the higher-kinded type @R@
to leave space for staging lattices, now we instantiate @R@ to @Rep@ and still use powerset as an 
example to present its staged version.

@@

\subsection{Staged Abstract Semantics} 
We have seen how to obtain a staged concrete semantics
based on types, now we take the same approach to obtain a staged abstract semantics.
\todo{what are static, what are dynamic?}
The basic operations largely kept same as the unstaged version, except the types are changed to @Rep@.
Besides, when we update the environment, the identifier @x@ is known statically, but the map of environment
has type @Rep[Map[Ident,Addr]]@, so we apply @lift@ to @x@ to make it as a next-stage value.

@@

\subsection{Staged Fixpoint Iteration.} 
\todo{staged fixed point iteration with memoization}

@@

Jones-optimal specialization?
Futamura projection?
Lattice-theoretic Interpretation
